---
title: "Return-to-Play After Concussion"
format: typst
execute:
  message: false
  warning: false
---

```{r}
library(tidyverse)
library(brms)
library(tidybayes)
library(marginaleffects)
```

## Data

One row per concussion event. Athletes with no concussions are excluded — they
contribute no outcome to the model.

```{r}
rtp_data <- read_csv("data/concussion_data.csv", show_col_types = FALSE) |>
  filter(!is.na(injury_date)) |>
  mutate(
    sex  = factor(sex, levels = c("M", "F")),
    city = factor(city)
  )

glimpse(rtp_data)
```

## Model

Lognormal regression for `days_to_rtp`. On the log scale, effects are additive
and exponentiate to multiplicative shifts in median RTP days.

Fixed effects:

- `scat5_initial` — injury severity drives longer recovery
- `sex` — female athletes recover more slowly at equivalent severity
- `concussion_num` — each repeat concussion extends recovery

Random effect:

- `(1 | city)` — training center captures unexplained protocol variation across
  the 8 centers

```{r}
#| output: false
#|
fit <- brm(
  days_to_rtp ~ scat5_initial + sex + concussion_num + (1 | city),
  family = lognormal(),
  data = rtp_data,
  chains = 4,
  cores = 4,
  backend = "cmdstan"
)
```

```{r}
summary(fit)
```

```{r}
pp_check(fit)
```

## Fixed Effects

Exponentiated coefficients are multiplicative factors on median RTP days. A
coefficient of 1.22 means 22% longer recovery.

```{r}
fit |>
  as_draws_df() |>
  select(b_Intercept, b_scat5_initial, b_sexF, b_concussion_num) |>
  pivot_longer(everything()) |>
  mutate(value = exp(value)) |>
  summarise(
    median = median(value),
    q2.5   = quantile(value, 0.025),
    q97.5  = quantile(value, 0.975),
    .by = name
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  arrange(desc(median)) |>
  knitr::kable()
```

```{r}
fit |>
  as_draws_df() |>
  select(b_scat5_initial, b_sexF, b_concussion_num) |>
  pivot_longer(everything()) |>
  mutate(name = case_match(
    name,
    "b_scat5_initial"  ~ "SCAT5 Initial",
    "b_sexF"           ~ "Sex (Female)",
    "b_concussion_num" ~ "Concussion #"
  )) |>
  ggplot(aes(exp(value), name)) +
  stat_pointinterval() +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  labs(
    x = "Multiplicative effect on median RTP days",
    y = NULL,
    title = "Posterior estimates: fixed effects on return-to-play duration"
  )
```

## Sex Effect

Marginal effect of sex on predicted RTP days, averaging over the observed
distribution of severity and concussion history.

```{r}
avg_comparisons(fit, variables = "sex") |>
  tibble() |>
  select(contrast, estimate, conf.low, conf.high) |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  knitr::kable()
```

## Training Center Random Effects

Posterior estimates of each center's deviation from the population mean RTP,
on the log scale. Positive values indicate longer-than-average RTP at that
center after accounting for athlete severity and sex.

```{r}
fit |>
  spread_draws(r_city[city, ]) |>
  ungroup() |>
  summarise(
    median = median(r_city),
    q2.5   = quantile(r_city, 0.025),
    q97.5  = quantile(r_city, 0.975),
    .by = city
  ) |>
  arrange(desc(median)) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable()
```

```{r}
fit |>
  spread_draws(r_city[city, ]) |>
  ggplot(aes(r_city, reorder(city, r_city))) +
  stat_pointinterval() +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    x = "Deviation from population mean log(RTP)",
    y = NULL,
    title = "Training center random effects",
    subtitle = "Positive = longer RTP than average after adjusting for severity and sex"
  )
```
