---
title: "Part 1: Concussion Data Simulation"
format: typst
execute:
  echo: false
  message: false
  warning: false
---

```{r}
library(tidyverse)
library(lubridate)

con <- DBI::dbConnect(RSQLite::SQLite(), "data/athlete_data.db")

athletes         <- tbl(con, "athletes")         |> collect()
sport            <- tbl(con, "sport")            |> collect()
elite_experience <- tbl(con, "elite_experience") |> collect()

DBI::dbDisconnect(con)
```

```{r}
#' Effects simulated for Part 2 recovery:
#'   - Sex
#'   - Training center

set.seed(52723)

CONCUSSION_RATE <- 0.28

sport_params <- tribble(
  ~sport          , ~concussion_multiplier , ~severity_shift ,
  "Rugby"         , 2.00                   ,  8L             ,
  "Hockey"        , 1.60                   ,  5L             ,
  "Lacrosse"      , 1.10                   ,  2L             ,
  "Soccer"        , 0.90                   ,  0L             ,
  "Flag Football" , 0.80                   , -3L             ,
  "Basketball"    , 0.60                   , -5L             ,
  "Baseball"      , 0.45                   , -8L             ,
  "Softball"      , 0.45                   , -8L
)

# Adding sports multiplier effects
weighted_mean_multiplier <- sport_params |>
  inner_join(count(sport, sport), by = join_by(sport)) |>
  summarise(sum(concussion_multiplier * n) / sum(n)) |>
  pull()

sport_params <- sport_params |>
  mutate(concussion_multiplier = concussion_multiplier / weighted_mean_multiplier)

training_centers <- athletes |>
  distinct(city) |>
  mutate(center_re = rnorm(n(), mean = 0, sd = 0.22))

# Female random effect to capture in model
athlete_covariates <- athletes |>
  select(athlete_id, sex, city) |>
  inner_join(training_centers, by = join_by(city)) |>
  mutate(sex_re = if_else(sex == "F", 0.10, 0.0))

athlete_career_spans <- athletes |>
  mutate(birthdate = as.Date(birthdate)) |>
  mutate(career_start_year = year(birthdate) + 18L) |>
  inner_join(
    elite_experience |>
      summarise(max_elite_year = max(elite_year), .by = athlete_id),
    by = join_by(athlete_id)
  ) |>
  mutate(
    career_start = make_date(career_start_year, 1, 1),
    career_end   = make_date(career_start_year + max_elite_year - 1L, 11, 30)
  ) |>
  select(athlete_id, career_start, career_end)

athlete_games <- elite_experience |>
  summarise(total_games = sum(games_played), .by = athlete_id)

mean_total_games <- mean(athlete_games$total_games)
RATE_PER_GAME    <- CONCUSSION_RATE / mean_total_games

# random effects
mean_rtp_shift <- mean(athlete_covariates$sex_re + athlete_covariates$center_re)

RTP_INTERCEPT  <- log(7) - mean_rtp_shift

# Conncussions simulated on the aggregated level
concussion_counts <- athletes |>
  select(athlete_id) |>
  inner_join(sport, by = join_by(athlete_id)) |>
  inner_join(sport_params, by = join_by(sport)) |>
  inner_join(athlete_games, by = join_by(athlete_id)) |>
  mutate(
    n_concussions = rpois(n(), lambda = RATE_PER_GAME * total_games * concussion_multiplier)
  )

# Turn each concussion into an event
concussion_events <- concussion_counts |>
  filter(n_concussions > 0) |>
  inner_join(athlete_career_spans, by = join_by(athlete_id)) |>
  inner_join(athlete_covariates, by = join_by(athlete_id)) |>
  mutate(concussion_num = map(n_concussions, seq_len)) |>
  unnest(concussion_num) |>
  mutate(
    career_days   = as.integer(career_end - career_start),
    injury_date   = career_start + map_int(career_days, \(d) sample.int(d, 1L)),
    scat5_initial = as.integer(pmin(
      pmax(round(rgamma(n(), shape = 5, scale = 8) + 5 + severity_shift), 5L), 132L
    )),
    severity_z    = (scat5_initial - 40) / 20,
    days_to_rtp   = as.integer(pmax(ceiling(rlnorm(
      n(),
      meanlog = RTP_INTERCEPT + (0.25 * severity_z) + (0.35 * (concussion_num - 1)) + sex_re + center_re,
      sdlog   = 0.5
    )), 1L)),
    return_to_play_date = injury_date + days_to_rtp,
    scat5_return  = as.integer(pmin(
      pmax(round(rgamma(n(), shape = 2, scale = 2.5) + 0.1 * scat5_initial), 0L),
      scat5_initial
    ))
  ) |>
  select(
    athlete_id, concussion_num, injury_date, return_to_play_date,
    days_to_rtp, scat5_initial, scat5_return
  )

concussions <- concussion_counts |>
  select(athlete_id, sport, total_games) |>
  left_join(concussion_events, by = join_by(athlete_id)) |>
  left_join(athlete_covariates |> select(athlete_id, sex, city), by = join_by(athlete_id))

concussion_events_only <- concussions |> filter(!is.na(injury_date))

concussions |>
  left_join(athletes |> select(athlete_id, height_cm, birthdate), by = join_by(athlete_id)) |>
  select(
    athlete_id, sex, height_cm, birthdate, city,
    sport, total_games,
    concussion_num, injury_date, return_to_play_date, days_to_rtp,
    scat5_initial, scat5_return
  ) |>
  write_csv("data/concussion_data.csv")
```


## Overview

### Simulation Assumptions

- Athletes begin their elite career at age 18
- Concussion counts follow $"Poisson"(0.28)$ per athlete
- Days to return to play follow a log-normal distribution, median ≈ 9 days
- SCAT5 scores range 0–132 (22 symptoms × 0–6 scale); initial score mean ≈ 40, return score mean ≈ 5

### Simulation Added effects

Sport drives the number of concussions
- Each sport has a different multiplier effect to the number of concussions.
- These are arbitrary values that follow what someone with minimal expertise would intuitively expect.

Sex and Training Centers impact RTP
- Female athletes have ~22% longer median RTP at equivalent severity
- Training centers each carry a random effect ~ $Normal(0, 0.22^2)$ on the log RTP scale

## Summary

```{r}
tibble(
  Metric = c(
    "Total athletes",
    "Athletes with ≥1 concussion",
    "Total concussions",
    "Mean concussions per athlete",
    "Median days to return to play",
    "Median SCAT5 initial score",
    "Median SCAT5 return score"
  ),
  Value = c(
    as.character(nrow(athletes)),
    as.character(n_distinct(concussion_events_only$athlete_id)),
    as.character(nrow(concussion_events_only)),
    as.character(round(nrow(concussion_events_only) / nrow(athletes), 2)),
    as.character(median(concussion_events_only$days_to_rtp)),
    as.character(round(median(concussion_events_only$scat5_initial), 1)),
    as.character(round(median(concussion_events_only$scat5_return), 1))
  )
) |>
  knitr::kable()
```

## By Sport


```{r}
concussions |>
  summarise(
    Athletes               = n_distinct(athlete_id),
    Concussions            = sum(!is.na(injury_date)),
    Rate                   = round(Concussions / Athletes, 2),
    `Median SCAT5 Initial` = median(scat5_initial, na.rm = TRUE),
    `Median SCAT5 Return`  = median(scat5_return,  na.rm = TRUE),
    `Median Days to RTP`   = median(days_to_rtp,   na.rm = TRUE),
    .by = sport
  ) |>
  arrange(desc(Rate)) |>
  knitr::kable()
```

## By Sex

```{r}
concussions |>
  summarise(
    Athletes               = n_distinct(athlete_id),
    Concussions            = sum(!is.na(injury_date)),
    Rate                   = round(Concussions / Athletes, 2),
    `Median SCAT5 Initial` = median(scat5_initial, na.rm = TRUE),
    `Median SCAT5 Return`  = median(scat5_return,  na.rm = TRUE),
    `Median Days to RTP`   = median(days_to_rtp,   na.rm = TRUE),
    .by = sex
  ) |>
  arrange(sex) |>
  knitr::kable()
```

## By Training Center

```{r}
concussions |>
  summarise(
    Athletes               = n_distinct(athlete_id),
    Concussions            = sum(!is.na(injury_date)),
    Rate                   = round(Concussions / Athletes, 2),
    `Median SCAT5 Initial` = median(scat5_initial, na.rm = TRUE),
    `Median SCAT5 Return`  = median(scat5_return,  na.rm = TRUE),
    `Median Days to RTP`   = median(days_to_rtp,   na.rm = TRUE),
    .by = city
  ) |>
  arrange(desc(`Median Days to RTP`)) |>
  rename(`Training Center` = city) |>
  knitr::kable()
```
