---
title: "Return-to-Play After Concussion"
format: typst
execute:
  message: false
  warning: false
  echo: false
---

```{r}
library(tidyverse)
library(brms)
library(tidybayes)
library(marginaleffects)
library(patchwork)
```

## Executive Summary

This report examines how long athletes take to return to play (RTP) following a concussion. Across 200 athletes and 10,177 games, 50 athletes sustained a concussion — 63 events in total.

#### Key Findings

- Female athletes take longer to return to play than male athletes — approximately 1.4 days longer on average.
- Higher initial SCAT5 scores predict longer recovery: each 10-point increase adds roughly 1.6 days before clearance.
- Return-to-play timelines differ meaningfully across training centers, even after accounting for injury severity and sex. **This points to inconsistent clearance practices rather than differences in the athletes themselves**.

#### Recommendation

Training centers should adopt a shared, protocol-driven clearance process that explicitly incorporates initial SCAT5 score. Without a common standard, return-to-play decisions are being made inconsistently — exposing some athletes to premature clearance and others to unnecessarily prolonged restriction.


```{r}
#| output: false
rtp_data <- read_csv("data/concussion_data.csv", show_col_types = FALSE) |>
  filter(!is.na(injury_date)) |>
  mutate(
    sex = factor(sex, levels = c("M", "F")),
    city = factor(city)
  )

rtp_data <- rtp_data |>
  mutate(scat5_initial_z = scale(scat5_initial)[, 1])


rtp_data |> head()

```

## Model

Recovery time was modeled as a function of three athlete-level factors and one facility-level factor:

- **Initial injury severity** (SCAT5 score at time of concussion)
- **Sex** (female / male)
- **Concussion history** (number of prior concussions)
- **Training center** (to detect facility-level variation after adjusting for the above)

```{r}
#| output: false
fit_prior <- brm(
  days_to_rtp ~ scat5_initial_z + sex + concussion_num + (1 | city),
  family = lognormal(),
  data = rtp_data,
  prior = c(
    prior(normal(2, 0.5), class = Intercept), # ~8-9 days
    prior(normal(0, 0.5), class = b),
    prior(normal(0, 0.5), class = sd),
    prior(normal(0, 0.5), class = sigma)
  ),
  chains = 4,
  cores = 4,
  backend = "cmdstan",
  sample_prior = 'only',
  control = list(adapt_delta = 0.95),
  seed = 52723
)
```

```{r}
#| output: false

# I followed this workflow of prior predictive checks.
pp_check(fit_prior)
```

```{r}
#| output: false
fit <- update(fit_prior, sample_prior = "no")
```

## Factors Affecting Recovery Time

The chart below shows each factor's estimated effect on days to return to play. Points to the right of the dashed line indicate longer recovery; points to the left indicate shorter recovery. The horizontal bars represent uncertainty — wider bars mean the data are less conclusive about that factor.

```{r}
fit |>
  as_draws_df() |>
  select(b_scat5_initial_z, b_sexF, b_concussion_num) |>
  pivot_longer(everything()) |>
  mutate(name = case_match(
    name,
    "b_scat5_initial_z"  ~ "SCAT5 Initial",
    "b_sexF"           ~ "Sex (Female)",
    "b_concussion_num" ~ "Concussion #"
  )) |>
  ggplot(aes(exp(value), name)) +
  stat_pointinterval() +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  labs(
    x = "Relative effect on recovery time (values > 1 = longer recovery)",
    y = NULL,
    title = "Estimated effect of each factor on return-to-play duration"
  )
```

## Effect Sizes in Days

The estimates below translate the chart above into concrete differences in recovery time, averaged across all athletes in the sample.

#### Sex

```{r}
avg_comparisons(fit, variables = "sex") |>
  tibble() |>
  select(contrast, days_to_rtp = estimate, conf.low, conf.high) |>
  mutate(
    across(where(is.numeric), \(x) round(x, 1)),
    contrast = "Female vs. Male",
  ) |>
  knitr::kable()
```

Females average 1.4 more days to RTP than males, after averaging over the observed distribution of injury severity and concussion history.

#### SCAT5

```{r}
scale_factor <- 10 / sd(rtp_data$scat5_initial)

scat5_comp <- avg_comparisons(fit, variables = "scat5_initial_z") |>
  tibble() |>
  select(contrast, days_to_rtp = estimate, conf.low, conf.high) |>
  mutate(
    across(where(is.numeric), \(x) round(x * scale_factor, 1)),
    contrast = "+10 points (raw)"
  )

scat5_comp |> knitr::kable()
```

Each 10-point increase in initial SCAT5 score is associated with an additional 1.6 days to RTP.

## Variation Across Training Centers

After adjusting for athlete severity and sex, there is meaningful variation in recovery time across training centers. The chart below shows each center's deviation from the average — centers to the right of the dashed line are clearing athletes to return later than expected, and centers to the left are clearing athletes earlier.


```{r}
fit |>
  spread_draws(r_city[city, ]) |>
  ggplot(aes(r_city, reorder(city, r_city))) +
  stat_pointinterval() +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    x = "Deviation from average recovery time (adjusted)",
    y = NULL,
    title = "Recovery time by training center, adjusted for athlete severity and sex",
    subtitle = "Right of dashed line = longer than average recovery"
  )
```

The chart below shows the expected days to return to play for a female athlete at median injury severity, broken out by training center.

```{r}
fit |>
  epred_draws(
    newdata = datagrid(
      model = fit,
      city = unique(rtp_data$city),
      scat5_initial = median(rtp_data$scat5_initial_z),
      concussion_num = 1,
      sex = "F"
    )
  ) |>
  ggplot(aes(x = .epred, y = reorder(city, .epred))) +
  stat_pointinterval() +
  geom_vline(xintercept = median(rtp_data$days_to_rtp), linetype = "dashed") +
  labs(x = "Adjusted median days to RTP", y = NULL)
```

## Appendix: Methodology

### Why a Bayesian Approach

With only 63 concussion events in the sample, a standard model would produce unreliably wide estimates. A Bayesian approach addresses this by combining the observed data with reasonable prior expectations — for example, that most athletes return to play within 30 days. The result is a range of plausible outcomes rather than a single number, giving a more honest picture of what the data can and cannot tell us.

The chart below shows the model's starting expectations (blue) versus its updated estimates after seeing the data (red). The dashed line is the actual observed distribution. The red curves closely track the observed data, indicating a well-calibrated model.

```{r}
set.seed(527)
n_draws <- 50

tidy_yrep <- function(model, label, n) {
  posterior_predict(model, ndraws = n) |>
    as_tibble(.name_repair = ~ paste0("obs", seq_along(.))) |>
    mutate(draw = row_number(), source = label) |>
    pivot_longer(-c(draw, source), values_to = "days_to_rtp")
}

yrep_df <- bind_rows(
  tidy_yrep(fit_prior, "Prior predictive", n_draws),
  tidy_yrep(fit, "Posterior predictive", n_draws)
)

obs_df <- tibble(days_to_rtp = rtp_data$days_to_rtp)

yrep_prior <- yrep_df |> filter(source == "Prior predictive")
yrep_post <- yrep_df |> filter(source == "Posterior predictive")
library(ggtext)

title_str <- "<span style='color:#1e47dc93'>**Prior**</span> → <span style='color:#e34646'>**Posterior**</span> Predictive Update"
ggplot(mapping = aes(days_to_rtp, group = draw)) +
  geom_density(
    data = yrep_prior,
    color = "#1e47dc93",
    linewidth = 0.3,
    alpha = 0.3
  ) +
  geom_density(
    data = yrep_post,
    color = "#e34646",
    linewidth = 0.3,
    alpha = 0.6
  ) +
  geom_density(
    data = obs_df,
    aes(days_to_rtp, group = NULL),
    color = "black",
    linewidth = 1,
    linetype = 'dashed'
  ) +
  coord_cartesian(xlim = c(0, 100)) +
  labs(
    x = "Days to RTP",
    y = "Density",
    title = title_str,
    subtitle = ("Dashed line is the observed distribution")
  ) +
  theme(plot.title = element_markdown())
```
