---
title: "Return-to-Play After Concussion"
format: typst
execute:
  message: false
  warning: false
  echo: false
---

```{r}
library(tidyverse)
library(brms)
library(tidybayes)
library(marginaleffects)
library(patchwork)
```

## Executive Summary

The following analysis was done on athlete concussion data, with the intent to understand the number of days for players to return to play (RTP) following a concussion. The data included 200 athletes across 10,177 total games. In the sample there were 50 athletes that suffered a concussion with a total of 63 concussions.

### Key insights include:

- Females have higher associated days before RTP.
- A 10-point increase in the initial SCAT5 is associated with roughly a 1.6 day increase in days to RTP.
- **There is inconsistency across training facilities in days prior to RTP, after controlling for severity and sex**.

### Recommendation

Each training facility should standardize on days to RTP with a guideline on the initial SCAT5 score.

## Data

The table below shows a sample of the concussion records used in this analysis.

```{r}
rtp_data <- read_csv("data/concussion_data.csv", show_col_types = FALSE) |>
  filter(!is.na(injury_date)) |>
  mutate(
    sex = factor(sex, levels = c("M", "F")),
    city = factor(city)
  )

rtp_data <- rtp_data |>
  mutate(scat5_initial_z = scale(scat5_initial)[, 1])


rtp_data |> head() |> knitr::kable()

```

## Model

Days to RTP were modeled with a log-normal regression. Outcomes are normally distributed on the log scale, so the effects are additive there and exponentiate to multiplicative shifts in RTP days. The parameters included were as follows:

Fixed effects:

- `scat5_initial` — injury severity at onset of concussion
- `sex` — female / male
- `concussion_num` — how many concussions the athlete had prior to the event

Random effect:

- `(1 | city)` — training center

### Model Justification

Due to a small sample size of concussion events, a Bayesian general linear regression model was implemented. This allows incorporation of priors, and provides uncertainty with the small sample size of data.

Rather than single point estimates, Bayesian inference yields a full posterior distribution for each parameter, giving a more honest picture of uncertainty given the limited number of concussion events.

```{r}
#| output: false
fit_prior <- brm(
  days_to_rtp ~ scat5_initial_z + sex + concussion_num + (1 | city),
  family = lognormal(),
  data = rtp_data,
  prior = c(
    prior(normal(2, 0.5), class = Intercept), # ~8-9 days
    prior(normal(0, 0.5), class = b),
    prior(normal(0, 0.5), class = sd),
    prior(normal(0, 0.5), class = sigma)
  ),
  chains = 4,
  cores = 4,
  backend = "cmdstan",
  sample_prior = 'only',
  control = list(adapt_delta = 0.95),
  seed = 52723
)
```

```{r}
#| output: false

# I followed this workflow of prior predictive checks.
pp_check(fit_prior)
```

```{r}
#| output: false
fit <- update(fit_prior, sample_prior = "no")
```

Each line represents a draw from the predictive distribution. The prior encodes the expectation that RTP is typically around 9 days and rarely exceeds 30 — uncertain, but within a plausible range. The posterior then updates on the observed data, converging on a distribution that closely matches the observed RTP.

```{r}
set.seed(527)
n_draws <- 50

tidy_yrep <- function(model, label, n) {
  posterior_predict(model, ndraws = n) |>
    as_tibble(.name_repair = ~ paste0("obs", seq_along(.))) |>
    mutate(draw = row_number(), source = label) |>
    pivot_longer(-c(draw, source), values_to = "days_to_rtp")
}

yrep_df <- bind_rows(
  tidy_yrep(fit_prior, "Prior predictive", n_draws),
  tidy_yrep(fit, "Posterior predictive", n_draws)
)

obs_df <- tibble(days_to_rtp = rtp_data$days_to_rtp)

yrep_prior <- yrep_df |> filter(source == "Prior predictive")
yrep_post <- yrep_df |> filter(source == "Posterior predictive")
library(ggtext)

title_str <- "<span style='color:#1e47dc93'>**Prior**</span> → <span style='color:#e34646'>**Posterior**</span> Predictive Update"
ggplot(mapping = aes(days_to_rtp, group = draw)) +
  geom_density(
    data = yrep_prior,
    color = "#1e47dc93",
    linewidth = 0.3,
    alpha = 0.3
  ) +
  geom_density(
    data = yrep_post,
    color = "#e34646",
    linewidth = 0.3,
    alpha = 0.6
  ) +
  geom_density(
    data = obs_df,
    aes(days_to_rtp, group = NULL),
    color = "black",
    linewidth = 1,
    linetype = 'dashed'
  ) +
  coord_cartesian(xlim = c(0, 100)) +
  labs(
    x = "Days to RTP",
    y = "Density",
    title = title_str,
    subtitle = ("Dashed line is the observed distribution")
  ) +
  theme(plot.title = element_markdown())
```

## Fixed Effects

Exponentiated coefficients are multiplicative factors on RTP days. A
coefficient of 1.2 means 20% longer recovery.

The posterior distributions of each fixed effect are shown below.

```{r}
fit |>
  as_draws_df() |>
  select(b_scat5_initial_z, b_sexF, b_concussion_num) |>
  pivot_longer(everything()) |>
  mutate(name = case_match(
    name,
    "b_scat5_initial_z"  ~ "SCAT5 Initial",
    "b_sexF"           ~ "Sex (Female)",
    "b_concussion_num" ~ "Concussion #"
  )) |>
  ggplot(aes(exp(value), name)) +
  stat_pointinterval() +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  labs(
    x = "Multiplicative effect on median RTP days",
    y = NULL,
    title = "Posterior estimates: fixed effects on return-to-play duration"
  )
```

## Interpretations

The fixed effects above are on the log scale. Marginal effects translate these into interpretable differences in days, averaged over the observed distribution of the other covariates.

### Sex

```{r}
avg_comparisons(fit, variables = "sex") |>
  tibble() |>
  select(contrast, days_to_rtp = estimate, conf.low, conf.high) |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  knitr::kable()
```

Females average 1.4 more days to RTP than males, after averaging over the observed distribution of injury severity and concussion history.

### SCAT5

`scat5_initial` was standardized for modeling. The table rescales the effect back to the original score units.

```{r}
scale_factor <- 10 / sd(rtp_data$scat5_initial)

scat5_comp <- avg_comparisons(fit, variables = "scat5_initial_z") |>
  tibble() |>
  select(contrast, days_to_rtp = estimate, conf.low, conf.high) |>
  mutate(
    across(where(is.numeric), \(x) round(x * scale_factor, 1)),
    contrast = "+10 points (raw)"
  )

scat5_comp |> knitr::kable()
```

Each 10-point increase in initial SCAT5 score is associated with an additional 1.6 days to RTP.

## Training Center Random Effects

Posterior estimates of each center's deviation from the population mean RTP,
on the log scale. Positive values indicate longer-than-average RTP at that
center after accounting for athlete severity and sex.


```{r}
fit |>
  spread_draws(r_city[city, ]) |>
  ggplot(aes(r_city, reorder(city, r_city))) +
  stat_pointinterval() +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    x = "Deviation from population mean log(RTP)",
    y = NULL,
    title = "Training center random effects",
    subtitle = "Positive = longer RTP than average"
  )
```




The plot below shows the expected RTP for a female athlete at median injury severity, broken down by training center. This makes the random effects interpretable on the original day scale.

```{r}
fit |>
  epred_draws(
    newdata = datagrid(
      model = fit,
      city = unique(rtp_data$city),
      scat5_initial = median(rtp_data$scat5_initial_z),
      concussion_num = 1,
      sex = "F"
    )
  ) |>
  ggplot(aes(x = .epred, y = reorder(city, .epred))) +
  stat_pointinterval() +
  geom_vline(xintercept = median(rtp_data$days_to_rtp), linetype = "dashed") +
  labs(x = "Adjusted median days to RTP", y = NULL)
```

Comparing male and female athletes side by side shows the consistent sex difference across all centers.

```{r}
fit |>
  epred_draws(
    newdata = datagrid(
      model = fit,
      city = unique(rtp_data$city),
      scat5_initial_z = median(rtp_data$scat5_initial_z),
      concussion_num = 1,
      sex = c("M", "F")
    )
  ) |>
  ggplot(aes(x = .epred, y = reorder(city, .epred), color = sex)) +
  stat_pointinterval(position = position_dodge(width = 0.5)) +
  geom_vline(xintercept = median(rtp_data$days_to_rtp), linetype = "dashed") +
  labs(x = "Adjusted median days to RTP", y = NULL, color = "Sex")
```
